Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-02 introduction.tex
spell check
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@837 ccdd05d5-56c1-461a-bf98-85f6876523b4

% Outline:
%  - Low-mass inspirals are the most promising source of gravitational radiation for advanced \textsc{ligo}
% - Motivation for high throughput
%   - Point out challenges of advanced detectors: more templates, longer templates
% - Motivation for low latency
%   - Observational applications of near-realtime detection
%   - Mention S6 and VSR2, online running, EM followup
% - Layout of paper: description of method, procedure followed

@@

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short GRBs)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
GRBs~\cite{1999Natur.398..400A,2003Natur.422..284F,2006A&A...454L.119J,0004-637X-660-1-489,2008Natur.455..183R,2011A&A...528A..15G}
by telescopes with extremely rapid response or, in the case of GRB 080319B, by
pure serendipity, where several telescopes were observing a previous GRB in the
same field~\cite{2008Natur.455..183R}. Short GRBs, on the other hand, typically
fade too quickly to observe past the initial burst of gamma rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short GRBs. An optical counterpart would vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
bandlimited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with predictive detection of compact binary sources is
possible with current computing resources.  Assuming other aspects of
gravitational-wave observation latency can be reduced significantly, this
should allow the possibility for prompt alerts to be sent to the astronomical
community.

>>>>

% Outline:
%  - Low-mass inspirals are the most promising source of gravitational radiation for advanced \textsc{ligo}
% - Motivation for high throughput
%   - Point out challenges of advanced detectors: more templates, longer templates
% - Motivation for low latency
%   - Observational applications of near-real-time detection
%   - Mention S6 and VSR2, online running, EM followup
% - Layout of paper: description of method, procedure followed

@@

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short GRBs)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
GRBs~\cite{1999Natur.398..400A,2003Natur.422..284F,2006A&A...454L.119J,0004-637X-660-1-489,2008Natur.455..183R,2011A&A...528A..15G}
by telescopes with extremely rapid response or, in the case of GRB 080319B, by
pure serendipity, where several telescopes were observing a previous GRB in the
same field~\cite{2008Natur.455..183R}. Short GRBs, on the other hand, typically
fade too quickly to observe past the initial burst of gamma rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short GRBs. An optical counterpart would vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with predictive detection of compact binary sources is
possible with current computing resources.  Assuming other aspects of
gravitational-wave observation latency can be reduced significantly, this
should allow the possibility for prompt alerts to be sent to the astronomical
community.
